### 2025-04-26 paper/ecai2025/appendices/scenarios.tex
start handling some reviewer point: why we choose vmas and the scenarion, clearer intro

\section{Environments}\label{apx:scenarios}
\begin{figure}
    \centering
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion.png}
        \caption{Dispersion}
        \label{fig:dispersion}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/transport.png}
        \caption{Transport}
        \label{fig:transport}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/discovery.png}
        \caption{Discovery}
        \label{fig:discovery}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/sampling.png}
        \caption{Sampling}
        \label{fig:sampling}
    \end{subfigure}
    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}\vspace{0.5cm}
\end{figure}

In this section, we provide a detailed discussion of the environment and scenarios used to test the \fname{} algorithm. Specifically, we utilize four vectorized environments available as part of VMAS.

VMAS is a differentiable simulator designed for various multi-agent tasks. These tasks are built on top of a differentiable physics engine, resulting in differentiable transition functions for all tasks. However, the reward function is non-differentiable. To address this limitation, we created custom reward functions, where possible, to replicate the non-differentiable ones provided by VMAS. Consequently, some scenarios are compatible with all tested algorithms (PPO, SHAC, and \fname{}), while others are only applicable to PPO and \fname{}.

Furthermore, the observations produced by VMAS are often only partial representations of the full state. While in some cases these observations are sufficient to predict the next observations, in others they are not. In such cases, \fname{} is expected to face challenges, as the world model is trained using only partial information. In contrast, SHAC computes gradients using access to the full world state. We chose to retain this limitation to better reflect the behavior of \fname{} in more realistic environments.

>>>>

\section{Environments}\label{apx:scenarios}
\begin{table}[htbp]
  \centering
  \caption{Comparison of Multi-Agent Reinforcement Learning Environments/Frameworks}
  \label{tab:marl_comparison}
  % Optional: resizebox to fit width if needed
  \resizebox{\textwidth}{!}{% 
  \begin{tabular}{@{}l l l l l l l l@{}}
    \toprule % Use booktabs rule, or \hline\hline
    \textbf{Feature} & \textbf{VMAS} & \textbf{PettingZoo} & \textbf{SMAC} & \textbf{MA-MuJoCo} & \textbf{PyBullet} & \textbf{Neural MMO} & \textbf{Melting Pot} \\
    \midrule % Use booktabs rule, or \hline
    Key Focus & Fast, Vectorized & Standardized & StarCraft II & Multi-Agent & General 3D & Large-Scale Agent & Social Dilemmas \\
     & 2D Physics & MARL API & Micromanagement & Robotics (MuJoCo & Physics & Survival/Econ. & / Cooperation \\
     & (PyTorch) & (Collection) & & tasks) & Engine & & \\
    \midrule % Separator line
    Single/Multi & Primarily Multi & Primarily Multi & Multi Only & Primarily Multi & Both & Multi Only & Multi Only \\
     & (can do Single) & (API standard) & & (adaptations) & & (Massively) & \\
    \midrule % Separator line
    Differentiable & Yes & No & No & No & No (standard & No & No \\
     & & & & & use) & & \\
    \midrule % Separator line
    Generality & Medium & Very High (via & Low (Specific game & Medium (MuJoCo & High & Medium (Large- & Medium (Social \\
     & (Physics-based, & wrapped envs) & benchmark) & robotics) & (General & scale agent focus) & dilemmas focus) \\
     & PyTorch) & & & & physics) & & \\
    \bottomrule % Use booktabs rule, or \hline
  \end{tabular}
  } % Closing bracket for resizebox, if used
\end{table}

\begin{figure}
    \centering
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion.png}
        \caption{Dispersion}
        \label{fig:dispersion}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/transport.png}
        \caption{Transport}
        \label{fig:transport}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/discovery.png}
        \caption{Discovery}
        \label{fig:discovery}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/sampling.png}
        \caption{Sampling}
        \label{fig:sampling}
    \end{subfigure}
    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}\vspace{0.5cm}
\end{figure}

In this section, we provide a detailed overview of the environments used to evaluate the \fname{} algorithm. For our experiments, we selected four vectorized scenarios from VMAS (Vectorized Multi-Agent Simulator), which offers an ideal balance of features for our research objectives.

VMAS stands out among alternative multi-agent reinforcement learning frameworks such as PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, PyBullet~\cite{DBLP:journals/corr/abs-1710-06347}, Neural MMO~\cite{DBLP:conf/aaai/DiehlB21}, and Melting Pot~\cite{DBLP:conf/icml/GuptaK0G20}. As summarized in Table~\ref{tab:marl_comparison}, VMAS provides several key advantages for our work:

\begin{compactitem}
    \item \textbf{Multi-Agent Framework}: VMAS is specifically designed for multi-agent scenarios, making it appropriate for our experimental setup.
    \item \textbf{Differentiable Physics}: Unlike most alternatives, VMAS offers a differentiable physics engine that enables gradient-based optimization techniques.
    \item \textbf{Vectorized Implementation}: The framework supports parallel execution of multiple environments, significantly improving training efficiency.
    \item \textbf{Physics-based Simulation}: VMAS provides realistic modeling of agent interactions and dynamics through its physics engine.
    \item \textbf{Open Source Accessibility}: The codebase is fully accessible, allowing necessary modifications for our experiments.
\end{compactitem}

While VMAS provides differentiable transition functions across all tasks, its reward functions are generally non-differentiable. To address this limitation, we implemented custom differentiable reward functions where possible to replicate the original VMAS behavior. Consequently, some scenarios are compatible with all tested algorithms (PPO, SHAC, and \fname{}), while others can only be used with PPO and \fname{}.

An additional consideration is that VMAS often provides agents with partial observations rather than complete state information. In some scenarios, these observations contain sufficient information to predict subsequent states, while in others they do not. This partial observability presents a particular challenge for \fname{}, as its world model must operate with incomplete information. In contrast, SHAC computes gradients with access to the full world state. We deliberately maintained these observability constraints to assess \fname{}'s performance under more realistic conditions.

Figure~\ref{fig:scenarios} illustrates the four VMAS environments used in our experiments, which are detailed in the following subsections.

### 2025-04-26 paper/ecai2025/sects/experiments.tex
first pass on spacing
start handling some reviewer point: why we choose vmas and the scenarion, clearer intro

\begin{figure*}[!t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Norm of policy gradients on the Transport environment across 5000 epochs. The raise in norm for SHAC (epoch 2700) corresponds with occurrence of generalization, where collision between agents and the packages become frequent. Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
\end{figure}

\section{Experiments}\label{sect:experiments}

In this section, we test \fname{} with PPO and SHAC for a set of multi-agent scenarios and architectures. An ablation study when replacing only the transition function (meaning that the reward function is assumed differentiable) is also conducted in \Cref{apx:ablation}.

\paragraph{Scenarios}
As mentioned previously, we choose a few multi-agent scenarios from the VMAS package to test \fname{}, SHAC and PPO\@. In particular, we choose the scenarios Dispersion, Discovery, Transport, and Sampling. For details regarding each scenario, see \Cref{apx:scenarios}. Dispersion and Discovery have non-differentiable reward functions, thus SHAC is not applicable. Transport and Sampling have differentiable reward functions, thus SHAC is applicable. For Transport and Dispersion, the observation space is complete wrt.\ the state. For Discovery and Sampling, the observation space is incomplete wrt.\ the state. In these scenario, \fname{} is slightly disadvantaged as it does not have access to the full state of the environment compared to SHAC\@.

\begin{table*}[!t]
    \centering
    \input tables/max.tex
    \caption{Normalized rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
\end{table*}

>>>>

\begin{figure*}[t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Norm of policy gradients on the Transport environment across 5000 epochs. The raise in norm for SHAC (epoch 2700) corresponds with occurrence of generalization, where collision between agents and the packages become frequent. Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
    \vspace{0.5cm}
\end{figure}
\section{Experiments}\label{sect:experiments}

In this section, we test \fname{} with PPO and SHAC for a set of multi-agent scenarios and architectures. An ablation study when replacing only the transition function (SHAC+,meaning that the reward function is assumed differentiable) is also conducted in \Cref{apx:ablation}.

\paragraph{Scenarios}
\begin{figure*}
    \centering
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion.png}
        \caption{Dispersion}
        \label{fig:dispersion}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/transport.png}
        \caption{Transport}
        \label{fig:transport}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/discovery.png}
        \caption{Discovery}
        \label{fig:discovery}
    \end{subfigure}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/sampling.png}
        \caption{Sampling}
        \label{fig:sampling}
    \end{subfigure}
    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}\vspace{0.5cm}
\end{figure*}
We evaluate our algorithms on several multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22}, which we selected over alternatives like PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, and others due to its key advantages: differentiable physics (enabling gradient-based optimization), multi-agent design, vectorized implementation (for efficient parallel training), realistic physics-based simulation, and open-source accessibility.
For more details about the choice of environments, see \Cref{apx:scenarios}.
We selected four distinct scenarios: Dispersion, Discovery, Transport, and Sampling, specifically for their cooperative nature and varying levels of required coordination (see \Cref{fig:scenarios} and \Cref{apx:scenarios} for details). These scenarios can be categorized along two dimensions:
\begin{compactitem}
    \item \textbf{Reward differentiability:} Dispersion and Discovery feature non-differentiable reward functions (making SHAC inapplicable), while Transport and Sampling have differentiable rewards (allowing for SHAC implementation).
    \item \textbf{Observation completeness:} Transport and Dispersion provide agents with complete observation spaces relative to the environment state. In contrast, Discovery and Sampling have incomplete observation spaces, which presents an additional challenge for \fname{} compared to SHAC, as \fname{} lacks access to the complete environmental state.
\end{compactitem}

This selection of scenarios enables a thorough evaluation of how each algorithm handles different types of multi-agent cooperation challenges.

\begin{table*}[t]
    \centering
    \input tables/max.tex
    \caption{Normalized rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
\end{table*}
